Standartox: A tool to derive exposure endpoints from chemical test results

### 2019-10-18 article/parts/02_methods.tex
Updates from Overleaf

Standartox consists of three software parts which are explained in detail below. Firstly a data acquisition and preparation processing pipeline (Fig. \ref{fig:pipeline}) downloads quarterly released new versions of the EPA ECOTOX data base and performs several preparation steps in R \citep{r_core_team_r_2017}, to then store the outcome in a PostgreSQL <CITE> data base. This constitutes the basis upon which the web application and and application programming interface (API) together with a R package are accessing the data.

@@

In order to build Standartox the EPA ECOTOX data is downloaded and subsequently built into a PostgreSQL data base. From there, the data is processed, including validity checks, removal of special characters, unit harmonizations and selection of relevant variables. Chemical abstract service (CAS) numbers, unique chemical identifiers together with taxon names are then used to query additional information from other data bases (Table \ref{tab:data-base}) on chemicals and organisms respectively. Standartox itself is finally built and exported as a binary object to be used by the web application and the API.

\subsection*{Web application and application programming interface}

Once called, both the web application and the API load the binary Standartox file into memory and allow the user to interact with it.

\subsubsection*{Web application}
The interactive web application was built in R, using the shiny web application framework \citep{chang_shiny_2018} which runs with the help of a shiny server \citep{HOW-TO-CITE-SHINY-SERVER} on www.standartox.uni-landau.de. The app itstel lies in the file ~/standartox-app/app.R.

\subsubsection*{Application programming interface (API)}
The API is built in R with the plumber-package \citep{}

@@

\subsubsection*{R-package}

httr,
jsonlite,
fst,
data.table

\subsubsection*{Aggregation}
The \etoxbase{} application aggregates the test results according to chosen filters in a two step process. Firstly the filtered test results are aggregated by the CAS number, the chosen taxon and the selected test duration. Secondly, the returned data is then aggregated by the CAS number. The former can't be influenced by the user and calculates either the minimum or the median depending on the amount of results to aggregate (n <= 2: minimum, if n > 2: median). Thereof the second step calculates the minimum, the maximum, the median, the geometric mean, or the arithmetic mean as an aggregate.

\subsubsection*{Future of the project}
Since new versions of the EPA ECOTOX data base are published on a quarterly basis the data acquisition process is automated and new versions of the Standartox are scheduled to be released regularly.

The code for the web-application and the application programming interface are placed here: https://github.com/andschar/standartox-app

\begin{figure}
    \includestandalone[scale=0.75]{tikz/application}
    \caption{Application}
    \label{fig:app}
\end{figure}

\subsection*{The R package}

The code for the R-package standartox is placed here: https://github.com/andschar/standartox

@@

The processing pipeline code is stored here https://github.com/andschar/standartox-build, the code for the web application is stored under https://github.com/andschar/standartox-app and the R-pckage code is stored in https://github.com/andschar/standartox. The code can freely be accessed under the MIT License <CITATION>. Most of the code is written in R 3.6.1 and some parts in PostgreSQL 9.6.1. An exhaustive list of additional R-pakage used can be found in the supplementary material <REFERENCE-TO-SUPPL>.

\begin{table}[h!]
    \centering
    \begin{tabular}{|l|l|l|l|}
        \hline
        Category        & Package       & Version   & Reference                     \\ \hline
        processing      & RCurl         & 1.95-4.11 & \citep{lang_rcurl_2018}       \\ \hline
        processing      & data.table    & 1.12.0    & \citep{dowle_data.table_2018} \\ \hline
        application     & shiny         & 1.1.0     & \citep{chang_shiny_2018}      \\ \hline
    \end{tabular}
    \caption{List of R packages used in compiling the data.\newline{}
    TODO: Automate version input and name generation - whole table\newline{}
    TODO2: debug chang\_shiny\_2018 reference}
    \label{tab:rpackages}
\end{table}

\section{TODO}

    \item it is planed to extend this tool to other data bases if they provide new test data  (e.g. UBA ETOC)

>>>>

Standartox consists of three parts of software explained in detail below. Firstly a data acquisition and preparation processing pipeline (Fig. \ref{fig:pipeline}) downloads quarterly released new versions of the EPA ECOTOX data base and performs several preparation steps in R \citep{rcoreteam_language_2017}, to then store the outcome in a PostgreSQL9.6.1 data base. This constitutes the basis upon which the web application and and application programming interface (API) together with a R package are accessing the data.

@@

In order to build Standartox the EPA ECOTOX data is downloaded and subsequently built into a PostgreSQL data base. From there, the data is processed, including validity checks, removal of special characters, unit harmonizations and selection of relevant variables. Chemical abstract service (CAS) numbers, unique chemical identifiers and taxa names are then used to query additional information from other data bases (Table \ref{tab:data-base}) on chemicals and organisms respectively. Standartox itself is finally built and exported as a compressed serialized object to be then used by the web application and the API.

\subsection*{Web application \& Application Programming Interface (API)}

Once called, both the web application and the API load the compressed serialized Standartox file into memory and allow a client to interact with it. The client can then call the functions stx\_filter() and stx\_aggregate() to filter and aggregate the data according to specific parameters (Table \ref{tab:app-parameters}).

\subsubsection*{Web application}
The interactive web application was built in R, using the shiny web application framework \citep{chang_shiny_2018} which runs with the help of a shiny server \citep{HOW-TO-CITE-SHINY-SERVER} on www.standartox.uni-landau.de. The app itself lies in the file ~/standartox-app/app.R.

@@

\subsubsection*{Application programming interface (API)}
The API is built by using the plumber-package \citep{trestletechnologyllc_plumber_2018} which allows for the creation of REST APIs from R. Four API-endpoints ('/catalog', '/filter/rds', 'aggregate' and 'meta') that can be queried from the client side were created. The \textit{catalog} API-endpoint returns a JavaScript Object Notation (JSON) file containing a catalog of possible filter parameters to choose from. The \textit{filter-rds} returns the filtered test results as a compressed serialized file \citep{klik_fst_2019}, to reduce size and allow for fast user queries. The \textit{aggregate} API-endpoint returns a R function to the client allowing the client the aggregation of the filtered data. Lastly, the \textit{meta} API-endpoint returns a JSON file with a timestamp of the request and the used Standartox version. The API is designed to be used with the R-package standartox and therefore uses serialization methods specific to R (rds() from the base-package and fst() from the fst-package). The code for the the API lies in the file ~/standartox-app/api.R.

\subsubsection*{R-package}
The R-package standartox contains two functions \textit{stx\_catalog()} and \textit{stx\_query()}. The former accesses the \textit{catalog} API-endpoint and returns a R-list of possible filter parameters that can be used in \textit{stx\_query()}. The latter returns a R list of three tables (R data.frames) containing a filtered data set, an aggregated data set and a table with the meta information retrieved from the \textit{meta} API-endpoint.

\subsubsection*{Standartox Aggregation}
Standartox aggregates the test results according to chosen filters in a two step process. Firstly the filtered test results are aggregated by the CAS number, the chosen taxon and the selected test duration. Secondly, the returned data is then aggregated by the CAS number. The former can't be influenced by the user and calculates either the minimum or the median depending on the amount of results to aggregate (n <= 2: minimum, if n > 2: median). Thereof the second step calculates the minimum, the maximum, the median, the geometric mean, or the arithmetic mean as an aggregate.

@@

Standartox's code is distributed on three Github repositories below. All code can freely be accessed under the MIT License <CITATION>. Most of the code is written in R 3.6.1 and associated packages (compare Supplement: \ref{list:r-packages}). Some parts in PostgreSQL 9.6.1.

\begin{itemize}

\item \url{https://github.com/andschar/standartox-build} \newline
Contains the code for the data acquisition and processing pipeline.

\item \url{https://github.com/andschar/standartox-app} \newline
Contains the code for the web application and the API.

\item \url{https://github.com/andschar/standartox} \newline
Contains the code for the R-package.

\end{itemize}

### 2019-10-18 article/parts/04_technical_validation.tex

\begin{figure}
    \includegraphics[width=1\linewidth]{article/figures/Standartox_PPDB.png}
    \caption{
        Comparison between Standartox and PPDB values
    }
    \label{fig:standartox_ppdb_diff}
\end{figure}

In order to validate \standartox we compared resulting values with values from other data bases, such as the PPDB. Of 542 chemicals, only 16\% diverged from the PPDB values by a factor greater than 10, supporting plausibility of the \standartox results \ref{fig:standartox_ppdb_diff}.

>>>>

\begin{figure}
    \includegraphics[width=1\linewidth]{article/figures/.png}
    \caption{
        Comparison between Standartox and PPDB values
    }
    \label{fig:standartox_ppdb_diff}
\end{figure}

\subsection{PPDB Comparsion}
The PPDB also offers effect values.

\subsection{Unit testing}

\subsection{Overall distribution of tests}
Heatmap: don't Include for now. Too large

In order to validate Standartox we compared resulting values with values from other data bases, such as the PPDB. Of 542 chemicals, only 16\% diverged from the PPDB values by a factor greater than 10, supporting plausibility of the \standartox results \ref{fig:standartox_ppdb_diff}.
